---
output:
  html_document: default
---
---
title: "analysis of weather eevnts affect on economics and public health"
output: html_notebook


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Download the raw data file and extract the data into a dataframe.Then convert to a data.table
```{r}
library("data.table")
library("ggplot2")

stormDF <- read.csv("repdata_data_StormData.csv.bz2")
```
# Converting data.frame to data.table
```{r}
stormDT <- as.data.table(stormDF)
```
### 2.2: Examining Column Names
```{r}
colnames(stormDT)
```
### 2.3: Data Subsetting

#Subset the dataset on the parameters of interest 

# Finding columns to remove
```{r}
cols2Remove <- colnames(stormDT[, !c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP")])
```
# Removing columns
```{r}
stormDT[, c(cols2Remove) := NULL]
```
# Only use data where fatalities or injuries occurred.  
```{r}
stormDT <- stormDT[(EVTYPE != "?" & 
(INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE"
  , "FATALITIES"
  , "INJURIES"
  , "PROPDMG"
  , "PROPDMGEXP"
  , "CROPDMG"
  , "CROPDMGEXP") ]

```
### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)

#Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.

# Change all damage exponents to uppercase.
```{r}
cols <- c("PROPDMGEXP", "CROPDMGEXP")
stormDT[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]
```
# Map property damage alphanumeric exponents to numeric values.
```{r}
propDmgKey <-  c("\"\"" = 10^0,
  "-" = 10^0, 
  "+" = 10^0,
  "0" = 10^0,
  "1" = 10^1,
  "2" = 10^2,
  "3" = 10^3,
  "4" = 10^4,
  "5" = 10^5,
  "6" = 10^6,
  "7" = 10^7,
  "8" = 10^8,
  "9" = 10^9,
  "H" = 10^2,
  "K" = 10^3,
  "M" = 10^6,
  "B" = 10^9)
```
# Map crop damage alphanumeric exponents to numeric values
```{r}
cropDmgKey <-  c("\"\"" = 10^0,
  "?" = 10^0, 
  "0" = 10^0,
  "K" = 10^3,
  "M" = 10^6,
  "B" = 10^9)
stormDT[, PROPDMGEXP := propDmgKey[as.character(stormDT[,PROPDMGEXP])]]
stormDT[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]
stormDT[, CROPDMGEXP := cropDmgKey[as.character(stormDT[,CROPDMGEXP])] ]
stormDT[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```
### 2.5: Making Economic Cost Columns
```{r}
stormDT <- stormDT[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, 
                       PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, 
                       CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```
### 2.6: Calcuating Total Property and Crop Cost

```{r}
totalCostDT <- stormDT[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
totalCostDT <- totalCostDT[order(-Total_Cost), ]
totalCostDT <- totalCostDT[1:10, ]
head(totalCostDT, 5)

```
### 2.7: Calcuating Total Fatalities and Injuries

```{r}
totalInjuriesDT <- stormDT[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
totalInjuriesDT <- totalInjuriesDT[order(-FATALITIES), ]
totalInjuriesDT <- totalInjuriesDT[1:10, ]
head(totalInjuriesDT, 5)

```
## 3: Results

### 3.1: Events that are Most Harmful to Population Health

#Melting data.table so that it is easier to put in bar graph format 
```{r}
bad_stuff <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "bad_thing")
head(bad_stuff, 5)

```

# Create chart
```{r}
healthChart <- ggplot(bad_stuff, aes(x=reorder(EVTYPE, -value), y=value))
```
# Plot data as bar chart
```{r}
healthChart = healthChart + geom_bar(stat="identity", aes(fill=bad_thing), position="dodge")
```
# Format y-axis scale and set y-axis label
```{r}
healthChart = healthChart + ylab("Frequency Count") 
```
# Set x-axis label
```{r}
healthChart = healthChart + xlab("Event Type") 
```
# Rotate x-axis tick labels 
```{r}
healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))
```
# Set chart title and center it
```{r}
healthChart = healthChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))

healthChart

```

### 3.2: Events that have the Greatest Economic Consequences

#Melting data.table so that it is easier to put in bar graph format 
```{r}
econ_consequences <- melt(totalCostDT, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_consequences, 5)
```
# Create chart
```{r}
econChart <- ggplot(econ_consequences, aes(x=reorder(EVTYPE, -value), y=value))
```
# Plot data as bar chart
```{r}
econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")
```
# Format y-axis scale and set y-axis label
```{r}
econChart = econChart + ylab("Cost (dollars)") 
```
# Set x-axis label
```{r}
econChart = econChart + xlab("Event Type") 
```
# Rotate x-axis tick labels 
```{r}
econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))
```
# Set chart title and center it
```{r}
econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") +
  theme(plot.title = element_text(hjust = 0.5))

econChart
```
